---
title: "Practical Examples of ggplot2 in Metaproteomics"
author: Suggestions to ningzhibin@gmail.com
date: Report generated @`r Sys.time()`
always_allow_html: TRUE
output:
  html_document:
    code_folding: show
    fig_width: 8
    fig_caption: TRUE
    toc: TRUE
    toc_depth: 4
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 4
    highlight: tango
    df_print: kable
    number_sections: TRUE
  word_document:
    toc: true
    toc_depth: 4
    highlight: tango
    df_print: kable
    smart: TRUE
---

```{=html}
<style type="text/css">
.main-container {
  max-width: 1400px;
  margin-left: auto;
  margin-right: auto;
}
</style>
```

# using geom_tile to plot the 2D ms data



```{r message=FALSE, warning=FALSE}

library(ggplot2)
library(tidyverse)
data <- rio::import("feature_data.csv")

head(data)


```

Any scatter plot function can visualize the 2d MS data easily, but not able to change the wide and height of the point. 
geom_tile has some hidden parameters to deal with that. The data mapping is 

Here we use a real feature data to demo. 


```{r message=FALSE, warning=FALSE}

plot(data[,c(2,3)], pch = 16, cex = 0.2)

```

With ggplot2(ggtile), it is easy to map the intensity into the color, with NA values set to grey

In ggtile, there are two hidden paras, height and width, most of the time, default settings will be good enough. However, if too many points or two few distant points, they will not/hardly visible

```{r message=FALSE, warning=FALSE}
data_small <- data.frame(X =sample(1:30,20), Y = sample(500:600,20), Z = sample(1:100000,100))
ggplot(data_small, aes(X, Y, fill= Z)) +geom_tile()

data_small <- data.frame(X =sample(1:3000,100), Y = sample(400:1600,100), Z = sample(1:100000,100))
ggplot(data_small, aes(X, Y, fill= Z)) +geom_tile()

```

Therefore, the hight and width needs to be properly set for complicated data. 

```{r message=FALSE, warning=FALSE}
 ggplot(data, aes(RT, MZ, fill= HM406CO)) +geom_tile(height = 10, width = 10)

```

```{r message=FALSE, warning=FALSE}
 ggplot(data, aes(RT, MZ, fill= HM406CO)) +geom_tile(height = 20, width = 50)

```





