---
title: "How to deal with function redundancy?"
author: Suggestions to techteam.metalab@gmail.com
date: Report generated @`r Sys.time()`
always_allow_html: TRUE
output:
  html_document:
    fig_width: 10
    fig_caption: TRUE
    toc: TRUE
    toc_depth: 4
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 4
    highlight: tango
    df_print: kable
    number_sections: TRUE
  word_document:
    toc: true
    toc_depth: 4
    highlight: tango
    df_print: kable
    smart: TRUE
---

<style type="text/css">
.main-container {
  max-width: 1400px;
  margin-left: auto;
  margin-right: auto;
}
</style>

# what is umap

[reference1](https://cran.r-project.org/web/packages/umap/vignettes/umap.html)

Uniform Manifold Approximation and Projection (UMAP) is an relatively newly developed algorithm for dimensional reduction. 
UMAP is good for large scale data, as well as traditional small dataset

there are 3 r packages for UMAP
umap and uwot, umapr

* [umapr](https://github.com/ropensci-archive/umapr) needs python. developed in 2018, discontinued and archived on github.
* [umap](https://cran.r-project.org/web/packages/umap/index.html) has two integrated function, one is python, but the default is r, no python needed. For more usage and details, check the [vignettes](https://cran.r-project.org/web/packages/umap/vignettes/umap.html). Easy to use. 
* [uwot](https://github.com/jlmelville/uwot) is still actively maintained. the author is co-author fo the second version of the umap publication.  it is a full implementation of the python version. 



## install the two package first

install.packages("umap")
install.packages("uwot")

```{r}
# prepare

library(vizier)
library(umap)
library(uwot)
library(ggpubr)
library(tidyverse)

```


# PCA of iris


```{r}

# PCA analysis
pca_iris <- stats::prcomp(iris[, -5], retx = TRUE, rank. = 2)

vizier::embed_plotly(pca_iris$x, iris$Species)
# df <- data.frame(pca_iris$x, Species = iris$Species)
# ggpubr::ggscatter(df, x = "PC1", y = "PC2", color = "Species") %>%  ggpubr::ggpar(legend = "right")


```



# using uwap::umap


```{r}
iris.data = iris[, grep("Sepal|Petal", colnames(iris))]

iris.umap = umap(iris.data)


vizier::embed_plotly(iris.umap$layout, iris$Species)
df <- data.frame(iris.umap$layout, Species = iris$Species)
ggpubr::ggscatter(df, x = "X1", y = "X2", color = "Species") %>%  ggpubr::ggpar(legend = "right")



```



# using uwot::umap

value is a matrix of optimized coordinates

```{r}
library(uwot)


# Non-numeric columns are ignored, so in a lot of cases you can pass a data
# frame directly to umap
iris_umap <- uwot::umap(iris, n_neighbors = 50, learning_rate = 0.5, init = "random")
vizier::embed_plotly(iris_umap, iris$Species)


```







```{r}

# Load mnist from somewhere, e.g.
# devtools::install_github("jlmelville/snedata")
# mnist <- snedata::download_mnist()
mnist_umap <- umap(mnist, n_neighbors = 15, min_dist = 0.001, verbose = TRUE)

# For high dimensional datasets (> 100-1000 columns) using PCA to reduce
# dimensionality is highly recommended to avoid the nearest neighbor search
# taking a long time. Keeping only 50 dimensions can speed up calculations
# without affecting the visualization much
mnist_umap <- umap(mnist, pca = 50)

# Use a specific number of threads
mnist_umap <- umap(mnist, n_neighbors = 15, min_dist = 0.001, verbose = TRUE, n_threads = 8)

# Use a different metric
mnist_umap_cosine <- umap(mnist, n_neighbors = 15, metric = "cosine", min_dist = 0.001, verbose = TRUE, n_threads = 8)

# If you are only interested in visualization, `fast_sgd = TRUE` gives a much faster optimization
mnist_umap_fast_sgd <- umap(mnist, n_neighbors = 15, metric = "cosine", min_dist = 0.001, verbose = TRUE, fast_sgd = TRUE)

# Supervised dimension reduction
mnist_umap_s <- umap(mnist, n_neighbors = 15, min_dist = 0.001, verbose = TRUE, n_threads = 8,
                     y = mnist$Label, target_weight = 0.5)

# Add new points to an existing embedding
mnist_train <- head(mnist, 60000)
mnist_test <- tail(mnist, 10000)

# You must set ret_model = TRUE to return extra data we need
# coordinates are in mnist_train_umap$embedding
mnist_train_umap <- umap(mnist_train, verbose = TRUE, ret_model = TRUE)
mnist_test_umap <- umap_transform(mnist_test, mnist_train_umap, verbose = TRUE)

# Save the nearest neighbor data
mnist_nn <- umap(mnist, ret_nn = TRUE)
# coordinates are now in mnist_nn$embedding

# Re-use the nearest neighor data and save a lot of time
mnist_nn_spca <- umap(mnist, nn_method = mnist_nn$nn, init = spca)

# No problem to have ret_nn = TRUE and ret_model = TRUE at the same time
# Or just use the ret_extra parameter:
mnist_nn_and_model <- umap(mnist, ret_extra = c("model", "nn"))

# You can also get to the input fuzzy graph as a sparse matrix via "fgraph"
mnist_with_fgraph <- umap(mnist, ret_extra = c("fgraph"))
# equivalent for lvish is to use "P" (input probability matrix):
mnist_with_P <- lvish(mnist, ret_extra = c("P"))

# Calculate Petal and Sepal neighbors separately (uses intersection of the resulting sets):
iris_umap <- umap(iris, metric = list("euclidean" = c("Sepal.Length", "Sepal.Width"),
                                      "euclidean" = c("Petal.Length", "Petal.Width")))
# Can also use individual factor columns
iris_umap <- umap(iris, metric = list("euclidean" = c("Sepal.Length", "Sepal.Width"),
                                      "euclidean" = c("Petal.Length", "Petal.Width"),
                                      "categorical" = "Species"))
```






# sessionInfo
```{r}
sessionInfo()
```

